---
title: "ordination-data-messing-around"
format: html
editor: visual
---

## 

Let's make a simple comparison of two communities

```{r}
site <- c("A", "B", "C", "D")
sugar_maples <- c(4, 1, 7, 5)

trees <- as.data.frame(cbind(site, sugar_maples))
trees$sugar_maples <- as.numeric(trees$sugar_maples)
```

Can we plot on one axis?

```{r}
library(tidyverse)

ggplot(trees, aes(sugar_maples, 0, color = site))+
  geom_point(size =3)+
  annotate("segment", x = 1, xend = 10, y = 0, yend = 0)+
  #annotate("segment", x = 1, xend = 1, y= 0.1, yend = 0.1)+
  annotate("segment", x = 10, xend = 10, y = 0.1, yend = 0.1)+
  scale_x_continuous(limits = c(0, 10), breaks = c(0,2,4,6,8))+
  scale_y_continuous(limits = c(-0.2, 0.2))+
  labs(x = "Number of sugar maple trees")+
  theme_bw()
  #theme(panel.background = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank()
        )


```

Now what if we had a second dimension? How about number of beech trees?

```{r}
trees$beech_trees <- c(5,3, 2, 4)
```

Now plot

```{r}
ggplot(trees, aes(sugar_maples, beech_trees, color = site))+
  geom_point(size =3)+
  #annotate("segment", x = 1, xend = 10, y = 0, yend = 0)+
  #annotate("segment", x = 1, xend = 1, y= 0.1, yend = 0.1)+
  #annotate("segment", x = 10, xend = 10, y = 0.1, yend = 0.1)+
  scale_x_continuous(limits = c(0, 10), breaks = c(2,4,6,8))+
  scale_y_continuous(limits = c(0, 10), breaks = c(2,4,6,8))+
  labs(x = "Number of sugar maple trees", y = "Number of beech trees")+
  theme_bw()


```

We can see when we look with two dimensions (number of species 1, number of species 2) that sites A and D are closer to one another than either is to either of the other 2 sites.

What if we have a 3rd species?

```{r}
trees$yellow_birch <- c(2, 5, 1, 2)
```

To plot in 3D we need `gg3d`

So, for each species that we add, we are adding an additional dimension to our analysis. This is helpful for differentiating among the sites, but we can't really plot more than 3 dimensions.

```{r}
library(gg3D)
ggplot(trees, aes(x = sugar_maples, y = beech_trees, z = yellow_birch, color = site))+
  theme_void()+
  axes_3D()+
  stat_3D(size = 2)
```

We can see that as we add dimensions, we start to see how the different sites cluster together or repel one another, but we have a problem: we can't really plot in \> 3 dimensions. We might have many more than 3 species, though, and we'd want data from those species.

The solution to this problem is ordination, in which, rather than plotting the actual values (number of trees) we build a distance matrix in which, depending on how we measure the distance, we get a single difference value for each pair of species.
